---
title: "Teaching Biology students to code smoothly with learnR and gradethis"
author:
- Engels Guyliann Engels, Numerical Ecology Department, Complexys and InforTec Institutes,
  University of Mons, Belgium
- Philippe Grosjean, Numerical Ecology Department, Complexys and InforTec Institutes,
  University of Mons, Belgium
date: "`r Sys.Date()`"
output:
  html_document:
    keep_md: yes
  pdf_document: default
bibliography: bibliography.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, results = "hide", message = FALSE, fig.path = "")

knitr::knit_hooks$set(
  plot = function(x, options) {
    hugoopts <- options$hugoopts
    paste0(
      "{", "{<figure src=", # the original code is simpler
      # but here I need to escape the shortcode!
      '"', x, '" ',
      if (!is.null(hugoopts)) {
        glue::glue_collapse(
          glue::glue('{names(hugoopts)}="{hugoopts}"'),
          sep = " "
        )
      },
      ">}}\n"
    )
  }
)

```

```{r}

SciViews::R
source("../R/learnr_functions.R")
pcloud_path <- "~/pCloud Drive/"

# 2020-2021 ----
pseudo_path <- paste0(pcloud_path, "sdd_2020-2021/data/")
#fs::dir_ls(pseudo_path)

learnr2020 <- read(paste0(pseudo_path, "learnr.csv")) %>.%
  mutate(., 
    user = paste0("id", user),
    course1 = stringr::str_extract(app, "^."),
    user_course = paste0(user, "_", course),
    user_course1 = paste0(user, "_", course1),
    app_label = paste0(app, "_", label))

## keep only the regular student only ----
user2020 <- read(paste0(pcloud_path, "Crypto Folder/sdd_users/users_2020-2021_q1.csv"))
staff <- paste0("id", 1:6)
us2020 <- filter(user2020, icflag %in% c("S-BIOG-006", "S-BIOG-015", "S-BIOG-025")) %>.%
  mutate(., 
    course1 = case_when(
      icflag == "S-BIOG-006" ~ "A",
      icflag == "S-BIOG-015" ~ "B",
      icflag == "S-BIOG-025" ~ "C"),
    user_course = paste0(ID, "_", icflag),
    user_course1 = paste0(ID, "_", course1))

stu_drop <- paste0("id", c(93, 68, 76, 164, 79, 69, 35, 33, 82, 153, 161))

learnr2020 %>.%
  filter(., !user %in% staff) %>.% # delete staff
  filter(., course %in% c("S-BIOG-006", "S-BIOG-015", "S-BIOG-025")) %>.%
  filter(., user_course1 %in% unique(us2020$user_course1)) %>.%
  filter(., !app %in% c(".A04a_graphs_adv", ".B01b_reg_poly", ".B01b_reg_multi", "A06Lb_recombinaison")) %>.%
  filter(., !user %in% stu_drop) %>.%
  mutate(., year = 2020) -> learnr2020

learnr2020 %>.%
  filter(., !label %in% stringr::str_subset(label, "noscore")) %>.%
  first_sequence(.) -> learnr_red2020

# 2019-2020 ----
library(DBI)
base_dir <- "~/pCloud Drive/Crypto Folder/sdd_data_backup/sdd_data_2019-2020/"
sdd2019 <- dbConnect(RSQLite::SQLite(), file.path(base_dir, "sdd.db"))
# Session that contains all event types
# dbListTables(sdd2019)
learnr2019 <- dbReadTable(sdd2019, "learnr")
dbDisconnect(sdd2019)

learnr2019$correct[learnr2019$correct == ""] <- NA

learnr2019 <- mutate(learnr2019,
  correct = as.logical(correct),
  date = lubridate::ymd_hms(date),
  course1 = stringr::str_extract(app, "^."),
  user_course = paste0(user, "_", course),
  user_course1 = paste0(user, "_", course1),
  app_label = paste0(app, "_", label)
)

learnr2019 <- mutate(learnr2019, key = oid(date))

## keep only the regular student only ----
user2019 <- read(paste0(pcloud_path, "Crypto Folder/sdd_users/users_2019-2020_q1.csv"))
us2019 <- filter(user2019, icflag %in% c("S-BIOG-006", "S-BIOG-015")) %>.%
  mutate(., 
    course1 = case_when(
      icflag == "S-BIOG-006" ~ "A",
      icflag == "S-BIOG-015" ~ "B"),
    user_course = paste0(ID, "_", icflag),
    user_course1 = paste0(ID, "_", course1))

stu_drop <- paste0("id", c(96, 144, 172, 173, 174, 175, 176, 159, 145, 179, 176, 100, 180))

learnr2019 %>.%
  filter(., !user %in% staff) %>.% # delete staff
  filter(., course %in% c("S-BIOG-006", "S-BIOG-015")) %>.%
  filter(., !app %in% c(
    ".00La_jde", "A..Lc_ch_examen", "A13Lc_examen", "A..Lb_m_examen", 
    "A10L._anova_test", "B..Lb_m_examen", "B..La_m_examen", "A..Lc_m_examen", "
    A11L._anova2_test", "A..La_examen")) %>.%
  filter(., user_course1 %in% unique(us2019$user_course1)) %>.%
  filter(., !user %in% stu_drop) %>.%
  mutate(., year = 2019) -> learnr2019

learnr2019 %>.%
  filter(., !label %in% stringr::str_subset(label, "comm")) %>.%
  first_sequence(.) -> learnr_red2019

learnr_comp <- bind_rows(select(learnr2019, names(learnr2020)), learnr2020)

learnr_red <- bind_rows(select(learnr_red2019, names(learnr_red2020)), learnr_red2020)
```

## Abstract

Students in a biology curriculum at the University of Mons learn to use R, RStudio, git and GitHub in their data science courses. The flipped classroom approach is used since 2018; students discover by themselves new concepts using online course material (<https://wp.sciviews.org>) at home. Then they apply the new concepts in the classroom on biological datasets in individual or group projects hosted on GitHub (<https://github.com/BioDataScience-Course>).

{Learnr} tutorials are used for a smooth transition between the theory at home and practice in the classroom. Students self-assess their conceptual understanding thanks to these tutorials with or without guided feedback using {gradethis}. {learnitdown} (<https://www.sciviews.org/learnitdown/>) is used to manage students identity and to record their activity in a MongoDB database.

Students in the first course are more perseverant when guided feedback is available, translating in a significant increase in the number of attempts per question. In the second course their perseverance was already high without the guided feedback, which exhibits thus a more limited effect.

Perception of the workload in the tutorials is quantified using a Raw Task load index (NASA-LTX questionnaire). This index could be used as one of the indicators to check the improvement of the tutorials in the future.

### Keywords

data science class, flipped classroom, learnr, gradethis, learnitdown

## Findings

Number of students , tutorials, questions and recorded events during the last two year are presented in the next table.

```{r, results='markup'}
learnr_comp %>.%
  group_by(., course1, year) %>.%
  summarise(., 
    user = length(unique(user)), app = length(unique(app)), 
    question = length(unique(app_label)), event = n()) -> tab

tab <- mutate(tab, course1 = case_when(
      course1 == "A" ~ "BioDataScience I",
      course1 == "B" ~ "BioDataScience II",
      course1 == "C" ~ "BioDataScience III")) 
  
knitr::kable(tab, col.names = c("Course", "Academic year","User", "App", "Question", "Event"))
```

The learnr tutorials analysed in the present study are available in GitHub:

-   BioDataScience1 (<https://github.com/BioDataScience-Course/BioDataScience1>)
-   BioDataScience2 (<https://github.com/BioDataScience-Course/BioDataScience2>)
-   BioDataScience3 (<https://github.com/BioDataScience-Course/BioDataScience3>)

```{r}

learnr_red %>.%
  group_by(., user, year, course, app_label, verb) %>.%
  summarise(., n = n()) -> learnr_verb

learnr_verb %>.%
  filter(., verb == "executed") %>.%
  group_by(., user, year, course) %>.%
  summarise(., tot = sum(n)) -> learnr_executed

learnr_verb %>.%
  filter(., verb == "submitted") %>.%
  group_by(., course, year)  %>.%
  summarise(., q_tot = length(unique(app_label))) -> learnr_submitted

learnr_participation <- left_join(learnr_executed, learnr_submitted, by = c("course", "year"))

learnr_participation <- mutate(learnr_participation, 
  ratio_question = tot/q_tot)

learnr_participation <- mutate(learnr_participation, 
  course_name = case_when(
    course == "S-BIOG-006" ~ "Biodatascience I",
    course == "S-BIOG-015" ~ "Biodatascience II",
    course == "S-BIOG-025" ~ "Biodatascience III"
    ),
  year_name = factor(case_when(
    year == 2019 ~ "without guided feedback",
    year == 2020 ~ "with guided feedback"
  ), levels = c("without guided feedback", "with guided feedback"))
  )

n_fun <- function(x){
  return(data.frame(y = max(x) * 1.1, label = paste0("n = ",length(x))))
}

chart(learnr_participation, ratio_question~course_name %fill=% year_name) +
  geom_hline(yintercept = c(5,10), alpha = 0.2) +
  geom_boxplot() +
  stat_summary(fun.data = n_fun, geom = "text", hjust = 0.5,
    position = position_dodge(0.75)) +
  labs(y = "Number of attempts / question", x = "Course", fill = "") +
    theme(legend.position = "top", legend.justification = "center")
```

The comparison with and without guided feedback does not show significant differences in number of attempts per exercise except for the Biodatascience I course (Wilcoxon independent test, W = 33, p-value = 2e-16). Guided feedbacks induce a behaviour change: students gave up sooner without them.

```{r}
learnr_participation %>.%
  filter(., course == "S-BIOG-006") %>.%
  wilcox.test(data = ., ratio_question ~ year_name)

learnr_participation %>.%
  filter(., course == "S-BIOG-015") %>.%
  wilcox.test(data = ., ratio_question ~ year)
```

### Perceived workload

The NASA-LTX indicator is composed of six questions on a Likert scale to quantify the perceived workload to complete a tutorial [@hart1988development]. The questions concern mental load, physical load, time pressure, expected success, effort required, and frustration experienced during the accomplishment of the task. The average value for the six questions constitutes a Raw Task Load indeX (RTLX) [@byers1989traditional].

```{r}
wo <- read(paste0(pseudo_path, "wooclap.csv"))

learnr_feeling <- function(df, apps, labels) {
  res <- dplyr::filter(df, app == apps & label == labels)
  
  c("(\\d):(Quelle a été l'importance de l'activité mentale et intellectuelle requise)", 
    "(\\d):(Quelle a été l'importance de l'activité physique requise)",
    "(\\d):(Quelle a été l'importance de la pression temporelle causée par la rapidité nécessitée pour l'accomplissement de la tâche)",
    "(\\d):Quelle réussite pensez-vous avoir eu dans l'accomplissement de la tâche",
    "(\\d):Quel degré d'effort avez-vous dû fournir pour exécuter la tâche demandée",
    "(\\d):Pendant l'exécution du travail vous êtes-vous senti découragé"
    ) %>.%
  purrr::map_dfc(., function(str, pattern) {stringr::str_match(str, pattern = pattern)[,2]}, str = res$value) -> vec
  
  names(vec) <- c("mental", "physical", "time_pressure", "performance", "effort", "frustration")
  
  vec %>.%
    map_dfc(.,as.numeric) -> vec1
  
  vec1$app <- apps
  vec1$label <- labels
  vec1$user <- res$user
  return(vec1)
}

c("A99Wa_perception", "B99Wa_perception", "C99Wb_perception:perception") %>%
  purrr::map_dfr(learnr_feeling, df = wo, label = "Q4") -> learnr_workload

learnr_workload$course <- stringr::str_extract(learnr_workload$app, "^.")

learnr_workload %>.%
  pivot_longer(.,cols = c(mental, physical, time_pressure, performance, effort, frustration),
  names_to = "category", values_to = "grade") -> workload

workload %>.%
  group_by(., user, app, course) %>.%
  summarise(., rtlx = 10*mean(grade)) -> workload_rtlx


chart(workload, grade ~ category %fill=% course) +
  geom_boxplot()
workload %>.%
  group_by(., user, app, course) %>.%
  summarise(., rtlx = 10*mean(grade)) -> workload_rtlx

workload_rtlx <- mutate(workload_rtlx, course_name = case_when(
  course == "A" ~ "Biodatascience I",
  course == "B" ~ "Biodatascience II",
  course == "C" ~ "Biodatascience III"
))


n_fun <- function(x){
  return(data.frame(y = max(x) * 1.1, label = paste0("n = ",length(x))))
}

chart(workload_rtlx, rtlx ~ course_name) +
  geom_boxplot(fill = "#00BFC4") +
  geom_jitter(alpha = 0.5, width = 0.1) +
  labs(y = "RTLX", x = "Course") +
  stat_summary(fun.data = n_fun, geom = "text", hjust = 0.5) +
  labs( y = "Raw Task Load indeX (2020-2021)")
```

Despite the increased difficulty of the exercises from one course to the other, RTLX indicator for Biodatascience III is significantly lower than for Biodatascience I (Tukey HSD, p-value = 0.023). We hypothesise that students are more used to the {learnr} tutorials as well as, R coding required to solve the exercises. Consequently, global workload is perceived as lighter. These RTLX measurements are meant to be a reference point for the following years: any future improvements in the tutorials should lead to similar or lower RTLX.

```{r}
workload_rtlx %>.%
  mutate(., course = as.factor(course)) -> workload_rtlx

anova. <- lm(data = workload_rtlx, rtlx ~ course)
anova(anova.)

#bartlett.test(data = workload_rtlx, rtlx ~ course)
#plot(anova., which = 2)

summary(anovaComp. <- confint(multcomp::glht(anova.,
  linfct = multcomp::mcp(course = "Tukey")))) # Add a second factor if you want
#.oma <- par(oma = c(0, 5.1, 0, 0)); plot(anovaComp.); par(.oma); rm(.oma)

```

The combination of {learnr}, {gradethis} and {learnitdown} provides a powerful tool to self-assess learning in data science. Students first exposed to {learnr} tutorials also benefit from additional guided feedback with {gradethis}. We propose to use the RTLX index as a mean to assess perceived difficulties and work required to complete these tutorials.


### **Availability of supporting source code and requirements**

-   Project home pages: <https://www.sciviews.org/learnitdown/>, <https://github.com/BioDataScience-Course/BioDataScience1>, <https://github.com/BioDataScience-Course/BioDataScience2>, <https://github.com/BioDataScience-Course/BioDataScience3>
-   Operating systems: Platform independent
-   Programming language: R
-   Other requirements: R 4.0.5 or higher, learnr 0.10.1.9008 or higher , gradethis, learnitdown 1.3.0 or higher

### Data availability

The data will be opened shortly.

### References

