---
title: 'Teaching Data Science in Biology: Supplemental Material'
author: "Philippe Grosjean & Guyliann Engels"
output:
  html_document:
    df_print: paged
  html_notebook: default
---

```{r setup}
SciViews::R
source("../R/sciviews_r_addons.R")
```

This document presents additional material to the manuscript "Teaching Data Science to Students in Biology using R, RStudio and Learnr: Analysis of Three Years Data", submitted for the special issue on "Data Science Education Research" of Foundations of Data Science.

## Multivariate analysis

In the manuscript, we use self-organizing map to create students' profiles. We have tested more conventional techniques like PCA, or hierarchical clustering, with less convincing results:

```{r}
sdd_metrics <- suppressMessages(read("../data/sdd_metrics.csv"))
metrics_nograde <- select(sdd_metrics, -user, -course, -state, -grade) # Only metrics
plot(correlation(metrics_nograde, method = "spearman"))
```

Distribution of the metrics:

```{r}
# Distribution of metrics
par(mar = c(4.1, 6.1, 2.1, 1.1)); boxplot(metrics_nograde,
  horizontal = TRUE, col = 'lightgray', las = 1, xlab = "Value")
```

```{r}
par(mar = c(4.1, 6.1, 2.1, 1.1)); boxplot(scale(metrics_nograde),
  horizontal = TRUE, col = 'lightgray', las = 1, xlab = "Value")
```

### PCA

```{r}
sdd_pca <- pca(metrics_nograde, scale = TRUE)
summary(sdd_pca)
```

```{r}
chart$scree(sdd_pca)
```

There are many PCs to consider here (8 to get 80%, 11 for 90%). This indicates that our metrics express many different features and are not easily summarized by a few axes. PCA is not suitable here.

With caution, we could look at first two planes, just to have an idea...

```{r}
chart$loadings(sdd_pca, choices = c(1, 2))
```

```{r}
chart$loadings(sdd_pca, choices = c(1, 3))
```

Just look if we differentiate de courses here... but interpret this very, very carefully!!!

```{r}
chart$score(sdd_pca, choices = c(1, 2), labels = sdd_metrics$course) +
  stat_ellipse()
```

```{r}
chart$score(sdd_pca, choices = c(1, 3), labels = sdd_metrics$course) +
  stat_ellipse()
```

We see, may be, a very slight evolution from A to B to C? But there are several extreme points.

### Groups

We use hierarchical clustering to group students into various profiles according to their activities.

```{r}
sdd_dist <- dissimilarity(metrics_nograde, method = "euclidean", scale = TRUE)
sdd_clust <- cluster(sdd_dist, method = "complete")
chart(sdd_clust) +
  geom_dendroline(h = 9, color = "red")
```

Not easy: the groups that are most easily separated contain very few students. Here again, the methods does not work very well.

```{r}
table(groups <- predict(sdd_clust, h = 9))
```

Most students are in one group!

```{r}
sdd_g <- augment(sdd_clust, sdd_metrics, h = 9)
table(sdd_g$.fitted, sdd_metrics$course)
```

No visible association between groups and courses.

```{r}
chart(data = sdd_g, grade ~ factor(.fitted)) +
  geom_boxplot()
```

Conclusion: PCA and hierarchical clustering do not produce convincing results here. This is why we used the Self-Organizing Map method instead.

### SOM

Self-Organizing Map produced for the publication and additional plots.

```{r}
library(kohonen)
set.seed(9433)
metrics_nograde %>.%
  scale(.) %>.%
  as.matrix(.) %>.%
  supersom(., grid = somgrid(3, 3, topo = "hexagonal"),
    rlen = 100L, mode = "online") ->
  sdd_som

plot(sdd_som, type = "changes")
```

```{r, eval=FALSE}
# Distribution of students according to courses and grades
set.seed(3566)
#set.seed(3749) The same set.seed used in the paper
par(mar = c(0, 0, 3.1, 3.1)); plot(sdd_som, type = "mapping", shape = "straight",
  col = scales::alpha(
    c(A = "red", B = "purple1", C = "blue")[as.factor(sdd_metrics$course)], 0.75),
  pch = 19, cex = (sdd_metrics$grade / 30) + 0.5, main = "")
#add.cluster.boundaries(sdd_som,  clustering = as.factor(sdd$course))
```

Same model, but different position of dots in the map, due to random parameter, in comparison to the publication. The graph below is the same as the previous one with the same layout as the graph in the paper.

```{r}
plot_som <- function(x, classif = NULL, main = "", labels = NULL, pchs = NULL,
  bgcol = NULL, keepMargins = FALSE, shape = c("round", "straight"),
  border = "black", margins = rep(0.6, 4), ...) {
  if (is.null(main))
    main <- "Mapping plot"
  #margins <- rep(0.6, 4)
  if (main != "")
    margins[3] <- margins[3] + 2
  if (!keepMargins) {
    opar <- par("mar")
    on.exit(par(mar = opar))
  }
  par(mar = margins)
  if (is.null(classif) & !is.null(x$unit.classif)) {
    classif <- x$unit.classif
  } else {
    if (is.list(classif) && !is.null(classif$unit.classif))
      classif <- classif$unit.classif
  }
  if (is.null(classif))
    stop("No mapping available")
  kohonen:::plot.somgrid(x$grid, ...)
  title.y <- max(x$grid$pts[, 2]) + 1.2
  if (title.y > par("usr")[4] - 0.2) {
    title(main)
  } else {
    text(mean(range(x$grid$pts[, 1])), title.y, main, adj = 0.5,
      cex = par("cex.main"), font = par("font.main"))
  }
  if (is.null(bgcol))
    bgcol <- "transparent"
  shape <- match.arg(shape)
  sym <- ifelse(shape == "round", "circle", ifelse(x$grid$topo ==
      "rectangular", "square", "hexagon"))
  switch(sym,
    circle = symbols(x$grid$pts[, 1], x$grid$pts[, 2],
      circles = rep(0.5, nrow(x$grid$pts)), inches = FALSE,
      add = TRUE, fg = border, bg = bgcol),
    hexagon = kohonen:::hexagons(x$grid$pts[,
      1], x$grid$pts[, 2], unitcell = 1, col = bgcol, border = border),
    square = symbols(x$grid$pts[, 1], x$grid$pts[, 2],
      squares = rep(1,
        nrow(x$grid$pts)), inches = FALSE, add = TRUE, fg = border,
      bg = bgcol))
  if (is.null(pchs))
    pchs <- 1
  if (is.null(labels) & !is.null(pchs))
    points(x$grid$pts[classif, 1] + rnorm(length(classif),
      0, 0.12), x$grid$pts[classif, 2] + rnorm(length(classif),
        0, 0.12), pch = pchs, ...)
  if (!is.null(labels))
    text(x$grid$pts[classif, 1] + rnorm(length(classif),
      0, 0.12), x$grid$pts[classif, 2] + rnorm(length(classif),
        0, 0.12), labels, ...)
  invisible()
}

cols <- scales::alpha(c(A = "#E69F00", B = "#56B4E9", C = "#009E73"), 0.75)
font_cex <- 0.7
plot_som(sdd_som, shape = "straight", col = cols[as.factor(sdd_metrics$course)],
  pch = 19, cex = (sdd_metrics$grade / 60) + 0.3, main = " ", keepMargins = TRUE,
  margins = rep(2.7, 4))
# Add numbers to cells
text((1:3) + 0.5, 3, 1:3, cex = font_cex)
text(1:3, 2.15, 4:6, cex = font_cex)
text((1:3) + 0.5, 1.3, 7:9, cex = font_cex)
legend(1.5, 3.65, legend = c("A", "B", "C"), col = cols, pch = 19,
  horiz = TRUE, bty = "n", title = "Course", cex = font_cex)
```

The Codes plot shows the importance of each variable on each cell. The number of variables is too large to describe each cell in detail.

```{r}
plot(sdd_som, type = "codes", codeRendering = "segments", shape = "straight")
```

Extract profile from the SOM analysis.

```{r}
sdd_metrics$profile <- sdd_som$unit.classif
```

How variables explain the distribution? In the manuscript, there is only a subset of these plots. Here are representations for all metrics.

```{r}
grays <- function(n, ...) {
  gray.colors(n, start = 1, end = 0, ...)
}
vars <- names(metrics_nograde)
var_names <- c(
  question     = "questions/module",
  `q_pub%`     = "% public questions",
  q_prod       = "contributions/question",
  `h_ok%`      = "correct H5P ex.",
  h_trials     = "trials/H5P ex.",
  `l_ok%`      = "correct learnr ex.",
  l_trials     = "trials/learnr ex.",
  l_hints      = "hints/learnr ex.",
  l_time       = "time/learnr ex.",
  i_commits    = "commits/ind. project",
  i_changes    = "contributions/ind. project",
  g_commits    = "commits/group project",
  g_changes    = "contributions/group project",
  `g_contrib%` = "% contribution to group projects",
  `done%`      = "work done",
  `intime%`    = "work done in time"
  )
for (var in vars) {
  par(mar = c(0, 0, 3.1, 0)); plot(sdd_som, type = "property",
    property = sdd_som$codes[[1]][, var],
    shape = "straight", main = "", palette.name = grays, heatkey = FALSE)
  title(var_names[var], cex.main = 1)
}
```

The plots in gray scale show how metrics distribute in the nine cells, from lowest value in white to highest value in black. The table below shows the values of each variable for each cell.

```{r}
codessom <- sdd_som[["codes"]][[1]]
row.names(codessom) <- c(7:9, 4:6, 1:3)

for(var in vars)
  colnames(codessom)[colnames(codessom) == var] <- var_names[names(var_names) == var] 

#codessom
codessom_table <- as.data.frame(t(codessom))
codessom_table[ , order(names(codessom_table))]
```

This single graph summarises the distribution of metrics in the nine cells, from the lowest value in white to the highest value in black, in grey scale.

```{r}

range01 <- function(x) {
    (x-min(x))/(max(x)-min(x))
}

codessom <- apply(codessom, 2, range01)
codessom <- rownames_to_column(as.data.frame(codessom), var = "gr")
codessom <- arrange(codessom, gr)

codessom %>.%
  pivot_longer(., `questions/module`:`work done in time`, names_to = "qu", values_to = "n") %>.%
  chart(., gr ~ qu %fill=% n) +
  geom_raster() +
  scale_fill_gradient(low = "white", high = "black") +
  theme(
    axis.text.x = element_text(angle = 80, hjust=1), 
    axis.title = element_blank())
```
