---
title: "Teaching Data Science in Biology: Supplemental Material"
author: "Philippe Grosjean & Guyliann Engels"
output: html_notebook
---

```{r setup}
SciViews::R
source("../R/sciviews_r_addons.R")
```

This document presents additional material to the manuscript "Teaching Data Science to Students in Biology using R, RStudio and Learnr: Analysis of Three Years Data", submitted for the special issue on "Data Science Education Research" of Foundations of Data Science.

## Multivariate analysis

In the manuscript, we use self-organizing map to create students' profiles. We have tested more conventional techniques like PCA, or hierarchical clustering, with less convincing results:

```{r}
sdd_metrics <- suppressMessages(read("../data/sdd_metrics.csv"))
metrics_nograde <- select(sdd_metrics, -user, -course, -state, -grade) # Only metrics
plot(correlation(metrics_nograde, method = "spearman"))
```

Distribution of the metrics:

```{r}
# Distribution of metrics
par(mar = c(4.1, 6.1, 2.1, 1.1)); boxplot(metrics_nograde,
  horizontal = TRUE, col = 'lightgray', las = 1, xlab = "Value")
```

```{r}
par(mar = c(4.1, 6.1, 2.1, 1.1)); boxplot(scale(metrics_nograde),
  horizontal = TRUE, col = 'lightgray', las = 1, xlab = "Value")
```

### PCA

```{r}
sdd_pca <- pca(metrics_nograde, scale = TRUE)
summary(sdd_pca)
```

```{r}
chart$scree(sdd_pca)
```

There are many PCs to consider here (8 to get 80%, 11 for 90%). This indicates that our metrics express many different features and are not easily summarized by a few axes. PCA is not suitable here.

With caution, we could look at first two planes, just to have an idea...

```{r}
chart$loadings(sdd_pca, choices = c(1, 2))
```

```{r}
chart$loadings(sdd_pca, choices = c(1, 3))
```

Just look if we differentiate de courses here... but interpret this very, very carefully!!!

```{r}
chart$score(sdd_pca, choices = c(1, 2), labels = sdd_metrics$course) +
  stat_ellipse()
```

```{r}
chart$score(sdd_pca, choices = c(1, 3), labels = sdd_metrics$course) +
  stat_ellipse()
```

We see, may be, a very slight evolution from A to B to C? But there are several extreme points.


### Groups

We use hierarchical clustering to group students into various profiles according to their activities.

```{r}
sdd_dist <- dissimilarity(metrics_nograde, method = "euclidean", scale = TRUE)
sdd_clust <- cluster(sdd_dist, method = "complete")
chart(sdd_clust) +
  geom_dendroline(h = 9, color = "red")
```

Not easy: the groups that are most easily separated contain very few students. Here again, the methods does not work very well.

```{r}
table(groups <- predict(sdd_clust, h = 9))
```

Most students are in one group!

```{r}
sdd_g <- augment(sdd_clust, sdd_metrics, h = 9)
table(sdd_g$.fitted, sdd_metrics$course)
```

No visible association between groups and courses.

```{r}
chart(data = sdd_g, grade ~ factor(.fitted)) +
  geom_boxplot()
```

Conclusion: PCA and hierarchical clustering do not produce convincing results here. This is why we used the Self-Organizing Map method instead.


### SOM

Self-Organizing Map produced for the publication and additional plots.

```{r}
library(kohonen)
set.seed(9433)
metrics_nograde %>.%
  scale(.) %>.%
  as.matrix(.) %>.%
  supersom(., grid = somgrid(3, 3, topo = "hexagonal"),
    rlen = 100L, mode = "online") ->
  sdd_som

plot(sdd_som, type = "changes")
```

```{r}
# Distribution of students according to courses and grades
set.seed(3566)
par(mar = c(0, 0, 3.1, 3.1)); plot(sdd_som, type = "mapping", shape = "straight",
  col = scales::alpha(
    c(A = "red", B = "purple1", C = "blue")[as.factor(sdd_metrics$course)], 0.75),
  pch = 19, cex = (sdd_metrics$grade / 30) + 0.5, main = "")
#add.cluster.boundaries(sdd_som,  clustering = as.factor(sdd$course))

```

Same model, but different position of dots in the map, due to random parameter, in comparison to the publication.

```{r}
plot(sdd_som, type = "codes", codeRendering = "segments")
```

Extract profile from the SOM analysis.

```{r}
sdd_metrics$profile <- sdd_som$unit.classif
```

How variables explain the distribution? In the manuscript, there is only a subset of these plots. Here are representations for all metrics.

```{r}
grays <- function(n, ...) {
  gray.colors(n, start = 1, end = 0, ...)
}
vars <- names(metrics_nograde)
var_names <- c(
  question     = "questions/module",
  `q_pub%`     = "% public questions",
  q_prod       = "contributions/question",
  `h_ok%`      = "correct H5P ex.",
  h_trials     = "trials/H5P ex.",
  `l_ok%`      = "correct learnr ex.",
  l_trials     = "trials/learnr ex.",
  l_hints      = "hints/learnr ex.",
  l_time       = "time/learnr ex.",
  i_commits    = "commits/ind. project",
  i_changes    = "contributions/ind. project",
  g_commits    = "commits/group project",
  g_changes    = "contributions/group project",
  `g_contrib%` = "% contribution to group projects",
  `done%`      = "work done",
  `intime%`    = "work done in time"
  )
for (var in vars) {
  par(mar = c(0, 0, 3.1, 0)); plot(sdd_som, type = "property",
    property = sdd_som$codes[[1]][, var],
    shape = "straight", main = "", palette.name = grays, heatkey = FALSE)
  title(var_names[var], cex.main = 1)
}
```

